Personalised Map Matched Imputation: an approach to impute missing data in smartphone location logs

### 2018-05-03 manuscript/apaManu.Rmd
last commit before preparation of research archive

---
title             : "Handling missing data in smartphone location logs"
shorttitle        : "Missing Data in Smartphone Location Logs"

@@

keywords          : "Missing Data, Measurement Bias, GPS, Human Mobility"
wordcount         : "2297"

@@

```{r load_packages, include = FALSE}
library("papaja")
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(ggthemes)
library(ggplot2)
library(knitr)
library(kableExtra)

@@

Why is human mobility important? How people move about in their environment affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010]. Therefore it is unsurprising that social scientists in numerous fields and even policymakers are interested in human mobility measures. The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@ipip]. In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g. @villanueva_association_2013]. Moreover, economists study geographic labour mobility extensively [e.g. @tatsiramos]. In fact, the European Commission considers labour mobility important enough to warrant its own directorate [@europeanComission]. The extensive use of measures like these across different domains strongly suggests that mobility metrics are linked to real world outcomes. Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive episode in bipolar depression patients [@palmius_detecting_2017]. Indeed, there is an argument to be made that perhaps psychologists have not been studying mobility enough. When studying behavioural differences within individuals behavioural scientists have often neglected the fact that individuals vary not only over time but also over space. To fully understand behaviour we must understand how behaviour can vary across environments.

Despite the importance of mobility measures, the majority of these measures are obtained through the use of questionnaires (such as the IPIP) or specifically through pen-and-paper travel diaries. Questionnaires and travel diaries are have well-known methodological flaws: they are burdensome to collect and rely on accurate self-reporting. The burdensomeness stems from the fact that participants must be explicitly asked to report on their movement patterns at frequent intervals due to forgetfulness. This makes the data expensive to collect and limits the extent of data collection in practice. In addition, the frequent reporting duties of the participant may bias the participants behaviour. Participant forgetfulness also limits the accuracy of self reported measures. There is clear evidence that participants are systematically biased when self-reporting mobility measures. For instance, participants under-report the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017]. These obstacles can be overcome by using objective data on human mobility.

Social scientists now have the unprecedented opportunity to easily obtain objective data on human mobility from smartphones. Before smartphones the only way to collect objective data on human mobility involved giving participants an expensive professional-grade location sensor and convincing them to take it with themselves at all times. @barnett_inferring_2016 points out that introducing a new device to the participant's life may bias their behaviour. Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016. Today millions of individuals carry smartphones with themselves every day and do not need to be encouraged to do so by researchers. Smartphones are equipped with a range of sensors that can be used to track the location of the device at all times. These smartphones can collect and store hundreds of location measurements a day. For instance,  Google Location History contains movement information on millions of users, often spanning years [@location_history_timeline_2017]. Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017]. Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millions of individuals spanning several years at little cost and without a significant burden of the participants.

@@

```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day. The first peak corresponds to the first journey from the user's home to a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre. Both journeys can be seen in Figure 1. The large jump between journey 5 and 6 is measurement error. The lower chart shows the accuracy over the course of the day. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual. Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.",out.width = '100%'}
include_graphics("img/accuracyLocShift.png")

@@

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! This issue is also visualised in Figure \@ref(fig:palmiusVme1). The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

@@

In the wider realm of spatiotemporal statistics there are numerous missing data imputation methods. These often come from climate or geological research and rely on spatiotemporal auto-correlations. For instance, the CUTOFF method estimates missing values by incorporating similar observed temporal information from the value’s nearest spatial neighbors [ @feng_cutoff:_2014 ]. The authors illustrate their example using rainfall data from gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values. The example provided relates to underground water data. Generally these models assume fixed measurement stations (such as rainfall gauging stations). For this reason they cannot be easily applied to missing mobility tracks without significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @Wu]. Importantly, none of them worked with secondary logs. Nonetheless we will detail what they did as informative examples. @palmius_detecting_2017 deal with the measurement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

@@

This method can be extended to imputing the data based on temporally, spatially or periodically close flights and pauses. In other words, for a given missing period, the individual's mobility can be estimated based on measured movements in that area, at that point in time or movements in the last 24 hours (*circadian proximity*).

On the other hand, @wu use what they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict future human movement. While the authors are concerned with prediction and not imputation, they devised a method called the Spatial Temporal Semantic (STS) algorithm which converts raw measurements to machine learning friendly discrete bins. Working with high-frequency measurements, @wu's method down-samples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways. This minimises measurement error and paves the way for applying machine learning methods to human mobility problems.

In this section we have argued that there is a lack of established practices to follow with respect to missing data in human mobility logs. Moreover, extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns in secondary logs. Finally we discussed in detail three approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016; @wu].

@@

Following @wu's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences. There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@_st-dbscan:_2007;]. We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. This implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.

@@

@wu's spatial-temporal-semantic (STS) feature extraction algorithm uses map matching as a ground truth to assign noisy measurements into discrete bins along roads. In other words, in addition to the measured data they also use a geographic database that contains information about the area in which the individual is (e.g. where precisely the roads are), and sort measurements into bins based on both the measurement and the geographic data base. For example, if an individual is measured as moving closely in parallel to a road A in an area where there is no other parallel road, @wu's method will assume that the individual is on the road A.

PMMI uses a similar logic, but without using any external geographic database. The key modification in PMMI is that whilst @wu uses a map from outside the persons location logs, we use the total location history of the individual to create a personalised map. This map is subsequently used to bin measurements. This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long. To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_{ab}$.

@@

```{r, measurementsPerDay,echo = FALSE,fig.cap="Missing data for the entire duration of the log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}
#fix width
include_graphics("img/missingdayBoaz5min.png")
my_citations <- cite_r(file = "r-references.bib")
```

For simplicity, we subsequently used a time period when the individual was living in the Netherlands. This leaves us with 156,000 measurements over a period of less than six months. Analyses as well as implementations and adaptions @palmius_detecting_2017's and @barnett_inferring_2016's model were performed using R and a multitude of other statistical packages `r my_citations`. All the code is available on a public repository [@mygithub].

@@

1. Evaluating the performance of the map building and assigning functions.
2. Comparing the performance of PMMI using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data (public transportation time-stamps).

@@

Before we can evaluate the accuracy of the imputations, it is essential to evaluate how well noise in the data has been cleaned. Otherwise we run the risk of over-fitting the model in the sense that we will measure the extent to which an imputation method can correctly impute measurement noise within $D_t$ instead of true location $G(t)$.

In order to evaluate the map building and binning we will first visually evaluate the paths and pause locations. A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy (e.g. @lcm). Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $\delta_{\text{dev}}$. With respect to the deviation distance $\delta_{\text{dev}}$, we expect:

@@

### Imputation algorithm performance

We will compare the performance of PMMI, @palmius_detecting_2017 and @barnett_inferring_2016. In addition, we will also compute a naive model, which simply imputes as the missing value the previous observed value. The naive model will serve as a baseline model. To compare the performance of these methods we will remove 25% of measured time intervals at random within a four week period. We will make our comparisons for intervals of 5 minutes, 1 hour and 1 day. In other words, we will remove 25% of time intervals at random, while varying the size of the time intervals removed.

@@

To compare all methods with each other we will compute a distance measure (how far was the removed location from the predicted location). For PMMI's imputations we will use a weighted mean for the distance measures whereby each 5 minute period is weighted equally. This is necessary because the other two model estimate far fewer values than PMMI. While @barnett_inferring_2016's and @palmius_detecting_2017's models estimate 12 measurements for each missing hour, PMMI estimates as many measurements as there are in the log. When individuals go to infrequently travelled locations they tend to use their phone's location services more than 100 more times, which leads to more measurements, and hence more imputed values. Hence, if the weights are not used PMMI is susceptible to error overestimates which are related to measurement frequency in a way that the other two methods are not.

In addition, other measures of interest for PMMI are accuracy (in what percentage of the cases was the appropriate cluster predicted), the *confidence* and the *distance expectation*. The confidence is the probability with which the model predicts the most likely cluster. For instance, if the model predicts the missing bin to be bin A with a probability of 0.9 then we can say it has high confidence in the prediction. Similarly, the distance expectation is the cross-product of the estimated probabilities that the individual is at any of all given clusters and the distances between the clusters to the true cluster. For example, if the true location of an individual is bin A (bin A is 10 meters away from bin B) and the model assigns a probability of 0.9 at bin A and 0.1 at bin B, then the distance expectation would be 1 meter.

Finally, we will take objective real world data and compare it to predicted values. We will use information from the Dutch public transportation card. The Dutch public transportation service provides users with time-stamped data of when and where they board, change lines or leave public transportation. To be able to make a comparison between models we will remove all measurements from within the 5 minute period that a time-stamped measurement is available. Then we will use each model to impute the location of the individual within that period.

@@

The selection of these parameters was more-or-less heuristically driven based on their on their ability to extract a meaningful personalised map. When selecting parameters there is a trade off is between bias and precision. This is because an increase in precision in the form of a higher for high density locations resolution comes at the expense of precision as assigning measurements to bins becomes more difficult. For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the expense of falsely categorising certain measurements to the wrong cluster. This is illustrated in Figure \@ref(fig:pauseLocUithof1).

@@

For the entire period examined period the we find a deviance of 40 meters and a median deviance of 15 meters. Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical $\frac{2}{3}$ value that is expected. Approximately 9% of values were not taken into account when creating the bins,given that their the accuracy $a$ exceeded $a_{\text{Path Lim}}$.

For the narrower period of March 2017 approximately 74% of the deviance values are within their corresponding accuracy values, which is close to the theoretical $\frac{2}{3}$ value that is expected. The raw unweighted mean and median deviance are 38 and 14 meters respectively. If we weight them such that each 5 minute interval is weighted equally to ease comparison with the other two methods, we find a mean deviance of 35 meters with a median of 12.8.

In comparison, Palmius' method has a median deviance $\delta_{\text{dev}}$ of 3 meters, with a mean of 115 due to high deviance outliers. On the other hand, Barnett's method has mean deviance $\delta_{\text{dev}}$ of 343 meters and a median deviance of 8 meters. Barnett's deviance is necessarily higher than Palmius', as they down-sample temporally (like Palmius) and subsequently aggregate into pauses and linear flights.

The key difference between temporal and spatial down-sampling is shown in Figure \@ref(fig:palmiusVme1). Temporal down-sampling is much more sensitive to noise in sparsely measured periods because it averages out values within five minute periods. Often there are only a few noisy measurements in those periods (see Figure \@ref(fig:longMeasurementsPerDay) ), which leads to a noise in the down-sampled values. Unsurprisingly, there is a positive relationship between deviance and the amount of measurements in each down-sampled interval. The fact that over 90% of deviance values are within accuracy (much higher than the expected theoretical 67%) confirms that temporal down-sampling is not sufficiently filtering out the noise.

```{r, palmiusVsMe1,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
#fix width
include_graphics("img/tempVsspace.png")
```

### Imputation evaluation

The @palmius_detecting_2017 model failed to impute 3% of all removed values for both the 5 minute and over 10% for the 1 hour tests. @palmius_detecting_2017' model failed to impute a single value for the day tests.
Similarly, @barnett_inferring_2016's method failed to impute over 11% of the missing values. PPMI made an imputation for all missing values. Table 1 shows the results of the distance metrics for each method.

@@

resultsDf <- data.frame( mean5min = c(82,43,269,426),
                         median5min = c(4,0,0,0),
                         mean1hr = c(345,497,908,1502),
                         median1hr = c(6,4,0,0),
                         mean1day = c(9273,NA,5757,14266),
                         median1day = c(12,NA,0,1288)
                         )

rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "PPMI","Naive Baseline")
colnames(resultsDf) <- c("Mean","Median",
                         "Mean","Median",
                         "Mean","Median")

# median 7 meters mean 60 not imputed 282 NA's Ian all data
options(knitr.kable.NA = '-')
apa_table(resultsDf, col_spanners = list(`Five minutes` = c(2, 3), `One Hour` = c(4, 5),`One Day` = c(6,7)), placement = "htb", digits = 0, caption = "Distance in meters between the removed time period and the imputed value.")
```

In terms of PPMI's accuracy in this period, the prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively. As a comparison, the naive model's accuracy ratings were 87%, 68% and 24% respectively. The distance expectation values were very similar to the distance scores, albeit approximately 5% higher. Confidence scores ranged from 0.006 to 1.

### Comparison with objective data

>>>>

---
title             : "Personalised Map Matched Imputation: imputing missing data from smartphone location logs"
shorttitle        : "Missing Data in Smartphone Location Logs"

@@

keywords          : "Missing Data, Measurement Bias, GPS, Human Mobility"
wordcount         : "8319"

@@

```{r load_packages, include = FALSE}
library("papaja")
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(ggthemes)
library(ggplot2)
library(knitr)
library(kableExtra)
library(sp)
library(leaflet)
library(raster)
library(geosphere)

@@

Why is human mobility important? Human mobility measures are quantified metrics of how people move about in their environment. Human mobility affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010]. The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@goldberg_international_2006]. At their root these questions are mobility measures. In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g. @villanueva_association_2013]. Economists are interested not only in where people live, but also where they go to work: geographic labour mobility is related to the income of individuals, the well-being of a country's economy and even informs the policy of bodies such as the European Comission [@tatsiramos_geographic_2009]. The extensive use of measures like these across different domains in social science strongly suggests that mobility metrics are linked to important real world outcomes. Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive episode in bipolar depression patients [@palmius_detecting_2017]. Indeed, there is an argument to be made that perhaps psychologists have not been studying mobility enough. When studying behavioural differences within individuals behavioural scientists have often neglected the fact that individuals vary not only over time but also over space. To fully understand behaviour we must understand how behaviour can vary across environments.

Despite the importance of mobility measures, the majority of these metrics are obtained through the use of questionnaires (such as the aforementioned personality questionnaire) or specifically through pen-and-paper travel diaries. As a thought experiment we encourage the reader to try to remember where exactly he or she was five weeks ago on Saturday at midday. Most people find this sort of task very difficult (if not impossible), which illustrates that questionnaires and travel diaries are have well-known methodological flaws. This method of collecting data is difficult and relies on accurate self-reporting. Travel diaries are burdensome to collect because participants must be explicitly asked to write down on their movement patterns at frequent intervals due to human forgetfulness. This makes the data collection expensive and limits the duration of data collection in practice. In addition, the frequent reporting duties of the participant may bias the participants behaviour. The limits of human cognition also limit the accuracy of self-reported measures. There is evidence that participants are systematically biased when self-reporting mobility measures. For instance, participants under-report the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017]. These obstacles can only be overcome by using objective data on human mobility.

Social scientists now have the unprecedented opportunity to easily obtain objective data on human mobility from smartphones. Before smartphones the only way to collect objective data on human mobility involved giving participants an expensive professional-grade location sensor and convincing them to take it with themselves at all times. @barnett_inferring_2016 points out that introducing a new device to the participant's life may bias their behaviour. Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016]. Today millions of individuals carry smartphones with themselves every day and do not need to be encouraged to do so by researchers. Smartphones are equipped with a range of sensors that can be used to track the location of the device at all times. These smartphones can collect and store hundreds of location measurements a day. For instance,  Google Location History contains movement information on millions of users, often spanning years [@location_history_timeline_2017]. Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017]. Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millions of individuals spanning several years at little cost and without a significant burden of the participants.

@@

```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day. The first peak corresponds to the first journey from the user's home to a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre. Both journeys can be seen in Figure 2. The large jump between journey 5 and 6 is measurement error. The lower chart shows the accuracy over the course of the day. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual. Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.",out.width = '100%'}
include_graphics("img/accuracyLocShift.png")

@@

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! This issue is also visualised in Figure \@ref(fig:palmiusVsMe1). The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

@@

In the wider realm of spatiotemporal statistics there are numerous missing data imputation methods. These often come from climate or geological research and rely on spatiotemporal auto-correlations. For instance, @feng_cutoff:_2014 estimate missing values by incorporating similar observed temporal information from the value’s nearest spatial neighbors. The authors illustrate their example using rainfall data from gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values. The example provided relates to underground water data. Generally these models assume fixed measurement stations (such as rainfall gauging stations). For this reason they cannot be easily applied to missing mobility tracks without significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @@wu_spatial-temporal-semantic_2017]. Importantly, none of them worked with secondary logs. Nonetheless we will detail what they did as informative examples. @palmius_detecting_2017 deal with the measurement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

@@

This method can be extended to imputing the data based on temporally, spatially or periodically close flights and pauses. In other words, for a given missing period, the individual's mobility can be estimated based on measured movements in that area, at that point in time or movements in the last 24 hours.

It is also worth mentioning that there is a body of work that incorporates road features and other background information to generate predictions or imputations. For instance, @liao_learning_2007 use hierarchical Markov models in combination with knowledge about the transportation system to bridge the gap between raw GPS sensor measurements and high level information such as a user's destination and mode of transportation. This can be then used to impute or predict missing steps. Along similar lines @wu_spatial-temporal-semantic_2017 use what they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict future human movement. While the authors are concerned with prediction and not imputation, they devised a method called the Spatial Temporal Semantic (STS) algorithm which converts raw measurements to machine learning friendly discrete bins. Working with high-frequency measurements, @wu_spatial-temporal-semantic_2017's method down-samples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways. This minimises measurement error and paves the way for applying machine learning methods to human mobility problems. Subsequently we will focus on methods which do not incorporate external information as they are more generally applicable.

In this section we have argued that there is a lack of established practices to follow with respect to missing data in human mobility logs. Moreover, extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns in secondary logs. Finally we discussed in detail three approaches which deal explicitly with mobility patterns from custom or secondary logs [@palmius_detecting_2017 ;@barnett_inferring_2016; @wu_spatial-temporal-semantic_2017].

@@

Following @wu_spatial-temporal-semantic_2017's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences. There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@_st-dbscan:_2007]. We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. This implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.

@@

@wu_spatial-temporal-semantic_2017's spatial-temporal-semantic (STS) feature extraction algorithm uses map matching as a ground truth to assign noisy measurements into discrete bins along roads. In other words, in addition to the measured data they also use a geographic database that contains information about the area in which the individual is (e.g. where precisely the roads are), and sort measurements into bins based on both the measurement and the geographic data base. For example, if an individual is measured as moving closely in parallel to a road A in an area where there is no other parallel road, @wu_spatial-temporal-semantic_2017's method will assume that the individual is on the road A.

PMMI uses a similar logic, but without using any external geographic database. The key modification in PMMI is that whilst @wu_spatial-temporal-semantic_2017 uses a map from outside the persons location logs, we use the total location history of the individual to create a personalised map. This map is subsequently used to bin measurements. This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long. To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_{ab}$.

@@

```{r, measurementsPerDay,echo = FALSE,fig.cap="Missing data for the entire duration of the log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. The entire log contains several long periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}
#fix width
include_graphics("img/missingdayBoaz5min.png")
my_citations <- cite_r(file = "r-references.bib", footnote = TRUE)
```

For simplicity, we subsequently used a time period when the individual was living in the Netherlands. This subset contains 156,000 measurements over a period of less than six months.

@palmius_detecting_2017's algorithm was implemented in R based on the original MATLAB code and pseudocode kindly provided by the author. @barnett_inferring_2016's method was slightly adapted in R to fit the the data structure from the original R code provided by the author. When executing their models we used the same parameters as the authors did. To represent @barnett_inferring_2016's model we used the variant where movements were sampled based on spatial proximity.

All analyses were performed using `r my_citations$r`.

`r my_citations$pkgs`

All the code is available on a public repository [@sobrado_my_2018].

@@

1. Evaluating the performance of the map building and assigning functions.
2. Comparing the performance of PMMI compared to alternative methods [[@palmius_detecting_2017; @barnett_inferring_2016]] and a baseline model using cross-validation.
3. Comparing the performance of PPMI compared to the aformentioned methods using objective ground-truth data (public transportation time-stamps locations).

@@

Before we can evaluate the accuracy of the imputations, it is essential to evaluate how well noise in the data has been smoothed. All methods smoothen noise differently in an attempt to extract true location $G(t)$ from measurements $D_t$. In the subsequent cross-validation step we will remove random smoothened blocks and compare the imputed value to the smoothened value. For this reason we want to ensure that the smoothened method actually does reflect the true location $G(t)$. Otherwise we run the risk of over-estimating the accuracy of a model in the sense that we will compute the extent to which an imputation method can correctly impute a smoothened value which has little to do with the true location $G(t)$.

In order to evaluate the map building and binning we will first visually evaluate the paths and pause locations. A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy [e.g. @brunsdon_path_2007]. Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $\delta_{\text{dev}}$. With respect to the deviation distance $\delta_{\text{dev}}$, we expect:

@@

### Imputation algorithm performance in cross-validation.

We will compare the performance of PMMI, @palmius_detecting_2017 and @barnett_inferring_2016. In addition, we will also compute two baseline models. The first one is a naive model, which simply imputes the previous observed value. The second one we will call the "home" model, which imputes that the individual is at home at all missing timepoints. Using these two models as baselines makes sense given the high degree of autocorrelation in the data and that individuals tend to spend most of their time at home.  To compare the performance of the aforementioned methods we will remove 25% of measured time intervals at random within a four week period. We will make our comparisons for intervals of 5 minutes, 1 hour and 1 day. In other words, we will remove 25% of time intervals at random, while varying the duration of the time intervals removed.

@@

To compare all methods with each other we will compute a distance measure (how far was the removed location from the predicted location) and the coverage. The coverage refers to the percentage of missing cases which were imputed. For PMMI's imputations we will use a weighted mean for the distance measures whereby each 5 minute period is weighted equally. This is necessary because the other two models downsample temporally to 5 minutes and because the frequency of measurements in a 5 minute period is correlated with imprecise measurements and visiting infrequently travelled locations. Hence weights must be used to avoid inflating error rates for PPMI in comparison to the other models.

In addition, other measures of interest for PMMI are *accuracy* (in what percentage of the cases was the appropriate cluster predicted), the *confidence* and the *distance expectation*. The confidence is the probability with which the model predicts the most likely cluster. For instance, if the model predicts the missing bin to be bin A with a probability of 0.9 then we can say this is a high confidence prediction. Similarly, the distance expectation is the cross-product of the estimated probabilities that the individual is at any of all given clusters and the distances between the clusters to the true cluster. For example, if the true location of an individual is bin A (bin A is 1000 meters away from bin B) and the model assigns a probability of 0.9 at bin A and 0.1 at bin B, then the distance expectation would be 100 meters.

### Imputation algorithm performance using external benchmarks.

We will compare the performance of PMMI, @palmius_detecting_2017, @barnett_inferring_2016 and the naive baseline model with regards to imputing real missing measurments. To do will use information from the Dutch public transportation card. The Dutch public transportation service provides users with time-stamped location data of when and where they board, change lines or leave public transportation. In this paper we used 97 such events in a single month.

To be able to make a comparison between models we will remove all measurements from within the 5 minute period that a time-stamped measurement is available. Then we will use each model to impute the location of the individual within that period and compare it to the true location.

@@

The model was tuned based on a visual inspection of the resulting personal map meeting the following criteria:

1. Plausible extraction of pause locations. For example, a pause in a field with no path leading to it is less plausible than a pause at a train station, airport, or office building.
2. A suitable extraction of paths. For instance, are ensuring that individual paths are complete and not merged unneccessarily in the case of closely parallel paths.
3. Less than a fifth of extracted clusters are assigned only one measurement. Unique clusters are unavoidable as people sometimes visit a location briefly and only once, nonetheless they pose problems for the classification algorithms.

When selecting parameters there is a trade off is between bias and precision. This is because an increase in precision (in the form of a higher resolution of locations location) comes at the expense of precision (assigning measurements to bins becomes more difficult). For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the risk of falsely categorising certain measurements to the wrong bin. This is illustrated in Figure \@ref(fig:pauseLocUithof1).

@@

For the entire period examined period the we find a deviance of 40 meters and a median deviance of 15 meters. Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical 67% value that is expected. Approximately 9% of values were not taken into account when creating the bins,given that their the accuracy $a$ exceeded $a_{\text{Path Lim}} = 300$.

In comparison, @palmius_detecting_2017's method has a median deviance $\delta_{\text{dev}}$ of 3 meters, with a mean of 115 due to high deviance outliers. On the other hand, @barnett_inferring_2016's method has mean deviance $\delta_{\text{dev}}$ of 343 meters and a median deviance of 8 meters. @barnett_inferring_2016's deviance is necessarily higher than @palmius_detecting_2017', as they down-sample temporally (like @palmius_detecting_2017) and subsequently aggregate into pauses and linear flights.

The key difference between temporal and spatial down-sampling is shown in Figure \@ref(fig:palmiusVme1). Temporal down-sampling is much more sensitive to noise in sparsely measured periods because it averages out values within five minute periods. Often there are only a few noisy measurements in those periods (see Figure \@ref(fig:longMeasurementsPerDay) ), which leads to a noise in the down-sampled values. Unsurprisingly, there is a positive relationship between deviance and the amount of measurements in each down-sampled interval. The fact that over 90% of deviance values are within accuracy (substantially higher than the expected theoretical 67%) confirms that temporal down-sampling is not sufficiently filtering out the noise.

```{r, palmiusVsMe1,echo = FALSE,fig.cap="The difference between temporal and spatial downsampling. The blue circles are raw measurments, the yellow circles are temporally downsampled locations. Spatially downsampled locations are in red. Due to measurement sparcity and inaccuracy many of the temporally downsampled locations are in unfeasable locations.",out.width = '100%'}
#fix width
include_graphics("img/tempVsspace.png")
```

### Imputation algorithm performance evaluation in cross-validation.

The first factor we must take into account is the coverage of each imputation model. The results show us that while PPMI and the naive baseline model impute all missing values, the other two models have difficulties doing so (Table 1). In particular, the models of @palmius_detecting_2017 and @barnett_inferring_2016's fail to impute an increasing amount of missing values with an increase in the duration of the missing period. @palmius_detecting_2017's method in particular failed to impute a single missing time period. The longer a log, the higher the probability that long time periods will be missing. Secondary logs are typically long. Hence missing data imputation methods designed to work with secondary logs must be capable of imputing longer time periods.

@@

resultsDf <- data.frame( mean5min = c(240,43,269,426,5599),
                         median5min = c(5,9,0,0,0),
                         imputed5min = c("97%","97%","100%","100%","100%"),
                         mean1hr = c(33,497,908,1502,5667),
                         median1hr = c(6,4,0,0,0),
                         imputed1hr = c("88%","89%","100%","100%","100%"),
                         mean1day = c(62,NA,5757,14266,5757),
                         median1day = c(6,NA,0,1288,0),
                         imputed1day = c("85%","0%","100%","100%","100%")
                         )

rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "PPMI","Naive Baseline","Home Baseline")
colnames(resultsDf) <- c("Mean","Median","Coverage",
                         "Mean","Median","Coverage",
                         "Mean","Median","Coverage")

options(knitr.kable.NA = '-')
apa_table(resultsDf, col_spanners = list(`Five minutes` = c(2, 4), `One Hour` = c(5, 7),`One Day` = c(8,10)), placement = "htb", digits = 0, caption = "Distance in meters between the removed time period and the imputed value.",landscape = T)
```
